---
title: "erin_demo"
author: "cam"
date: '2018-02-16'
output: html_document
---

# R Markdown Demo

This is a **demonstration** to show you how R Markdown works. Here is a helpful [link](https://www.rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf) to find out more about R Markdown.

## 1. Loading necessary packages
```{r, message = FALSE}
library(readr)
library(dplyr)
library(visreg)
```

{r,results='hide'} would let you hide the results

{r,echo = FALSE} to hide the code 

\$ to get just a plain dollar sign

## 2. Packages needed for R markdown
```{r}
library(rmarkdown)
library(knitr)
```
## 3. Setting working directory
```{r}
setwd("/Users/Cam/Desktop/Code/UGRU/jupyter_notebook")
```

## 4. We are going to use the dataset "ToothGrowth", which looks at the effect of VitC on tooth growth in Guinea Pigs. Information can be found here:
?ToothGrowth

#5. Get dimensions of dataframe
dim(ToothGrowth)

#6. Summarize data
summary(ToothGrowth)

#7. It looks like we need to change supplement type to a factor
ToothGrowth$supp <- as.factor(ToothGrowth$supp)

#8. Let's gather some basic descriptive stats. What is the mean and standard error for tooth length?
mean(ToothGrowth$len)
sqrt(var(ToothGrowth$len)/length(ToothGrowth$len))

#9. Lets look at the effect of dosage on tooth length
summary(lm1 <- lm(len ~ dose, data=ToothGrowth))

#10. Check diagnostics
par(mfrow=c(2,2))
plot(lm1)
par(mfrow=c(1,1))
hist(resid(lm1))

#11. Plot this relationship
visreg(lm1, "dose")

#12. Lets try another model to look at the effect of the type of supplement on tooth length
summary(lm2 <- lm(len ~ supp, data=ToothGrowth))

#13. Plot this relationship
visreg(lm2, "supp")

#14. How about an interaction between dose and supplement type?
summary(lm3 <- lm(len ~ dose*supp, data=ToothGrowth))

#15. Plot this relationship
visreg(lm3, "dose", by="supp")

